An upper bound to gas storage and delivery via pressure-swing adsorption in porous materials

### 2021-07-09 gas-adsorption/rsc-gas-ads-complete.tex
fix to latex
edit caption

\usepackage[color]{changebar}
\cbcolor{red}
\newcommand\startreferees[1]{\cbstart \noindent\textit{\color{red}%
[\textbf{To referees:} #1]}

}
\newcommand\donereferees{\cbend}

@@

\newcommand{\davidsays}[1]{{\color{red} [\blue{D:} \emph{#1}]}}
\newcommand{\jpsays}[1]{{\color{red} [\blue{Jordan:} \emph{#1}]}}
\newcommand{\corysays}[1]{{\color{red} [\blue{Cory:} \emph{#1}]}}

@@

\startreferees{Deliverable capacity defined below}
For a vehicle employing a porous material to store natural gas or hydrogen, the
(volumetric) \emph{deliverable capacity} of the gas is the primary
thermodynamic property that determines the driving
range on a given ``full'' tank of fuel~\cite{mason2014evaluating}. 
The adsorbed gas storage tank delivers the
gaseous fuel to the engine via an (assumed) isothermal pressure
swing~\cite{sircar2002pressure}. The deliverable capacity (see
Fig.~\ref{fig:delcap}) is the density of the gas in the material at the storage
pressure $\pfull$ minus the residual gas that remains adsorbed at the lowest
pressure $\pempty$ such that sufficient flow is maintained to feed the engine.
For commercial feasibility, the US Department of Energy (DOE) has set
deliverable capacity targets for both adsorbed methane and hydrogen storage and
delivery for vehicles. For methane, the Advanced Research Projects
Agency--Energy (ARPA-E) set a deliverable capacity target of 315\ L STP CH$_4$
per L of adsorbent at 298 K using a 65\,bar to 5.8\,bar pressure
swing~\cite{simon2015materials}. For hydrogen, the DOE set a series of
progressive targets at five year intervals, with the ultimate target of 0.05\,kg
H$_2$/L~\cite{h2targetsDOE} using a 100\ bar to 5\ bar pressure swing at a
minimum of -40 $^\circ$C~\cite{allendorf2018assessment}. Thus far, despite the
emergence of highly tunable materials with large surface areas, such as
metal-organic frameworks~\cite{furukawa2013chemistry} and covalent organic 
frameworks~\cite{diercks2017atom}, no porous material has met these deliverable capacity targets.

\donereferees

@@

\startreferees{Added discussion of screening studies}
Finally, a number of empirical studies have been performed that computationally
simulate a adsorption in a vast array of possible porous
structures~\cite{firlej2013understanding, goldsmith2013theoretical,
ahmed2019exceptional}.  Such studies can guide the development of new materials
by allowing researchers to identify commonalities among materials that have
desirable properties.  However, such an approach can only place a \emph{lower
bound} on the highest attainable deliverable capacity.

\donereferees

@@

\subsection{The real-substrate analog of $\V$ is $\gst$}
\label{sec:phi-is-delta-g}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{g_st_simple.pdf}
    \caption{Cartoon illustrating the definition of $\gst$. For a given adsorbent and gas, $\gst$=$\gst (\rho, T)$ is defined as the difference in chemical potential between the adsorbed gas system and bulk gas system when both exhibit the same density $\rho$ and temperature $T$.}
    \label{fig:delta-G-cartoon}
\end{figure}

@@

\begin{equation}
   \gst(\rho, T) \equiv
   %\frac{G_{\text{ads}}(\rho, T) - G_{\text{gas}}(\rho, T)}{N}\\
    \mu_{\text{ads}}(\rho, T) - \mu_{\text{gas}}(\rho, T).
  % \gst(\rho, T) &\equiv g_{\text{gas}}(\rho, T) - g_{\text{ads}}(\rho, T)
  \label{eq:g_st}
\end{equation}
The isosteric Gibbs free energy difference $\gst$ is the difference in molar
Gibbs free energy (equivalent to chemical potential) between the adsorbed gas system and
the bulk gas \emph{with the same density of gas molecules} and at the same temperature. The quantity $\gst$
does \emph{not} correspond to a change in the molar Gibbs free energy as a
molecule is adsorbed, which is zero under conditions of coexistence. The
quantity $\gst$ in a real substrate is a direct analog to $\V$ in our ideal
substrate because it is the chemical potential shift needed to impose on the
bulk gas in coexistence with the real substrate to achieve the same density as
in the substrate (compare with eqn.~\ref{eq:xi_vs_xi0}).
% Figure~\ref{fig:delta-G-cartoon} illustrates  $\gst$.
% David:  no longer illustrates the expt.
% A hypothetical experiment to measure $\gst$ is via a piston with a removable partition that separates a volume of free space from the same volume of substrate.
% Initialize the experiment with all gas in the free space and an empty substrate.
% Then, slowly (i) remove the partition and (ii) use the piston to push the gas into the substrate and eliminate the free space volume, in an isothermal process.
% $\gst$ is the ... \corysays{wut? energy it takes to do this or?}
%
% \davidsays{Okay, here is an improved experiment to measure $\gst$.  Two containers of
% equal volume are connected by a hose (with neglible volume).  One container holds
% the substrate, and the other is empty (apart from gas).    The altitude of one container
% is adjusted until the number of gas molecules is equal in each container.  The value
% of $\gst$ is the gravitational potential energy difference of gas molecules in the two
% boxes.}
%
Figure~\ref{fig:delta-G-cartoon} illustrates a hypothetical experiment to
measure $\gst$.  Two containers of equal volume are connected by a hose.  One
container holds the substrate, and the other is empty (apart from gas).    The
altitude of one container is adjusted until the number of gas molecules is equal
in each container.  The value of $\gst$ is the gravitational potential energy
difference of gas molecules in the two boxes.

% David: what is GST
Note that $\gst$ is a property
of both the substrate and the identity of the gas. Because real substrates
offer a \emph{non}-spatially uniform potential, $\gst(\rho, T)$ is a function
of $\rho$ and $T$, unlike our ideal, homogenous substrate where $\gst(\rho,
T)=\V$. Consequently, throughout this article, we show $\gst(\rho, T)$ for real
substrates at both conditions relevant to gas storage and delivery, $\pfull$
and $\pempty$.

@@

\startreferees{Flexible MOFs section added below}
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{figs/two-phase}
    \caption{Cartoon of a two-phase, collapsing material. The material can exist in two phases: an open phase in which gas molecules can adsorb and a closed phase that is collapsed and cannot fit any gas molecules. Take the closed phase as more stable, so that the material can collapse and expels residual gas.
    The pressure of gas at which the structural transition occurs depends on the gas-host interactions and relative stability of the phases. We wish for the structural transition to occur between $\pfull$ and $\pempty$.
    }
    \label{fig:flexible-cartoon}
\end{figure}

\section{Flexible two-phase model}
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{methane-298-flexible}
    \caption{Maximal possible deliverable capacity of methane in a two-phase, collapsing substrate at 298\ K as a function of the free energy difference 
      between open and closed phases. The experimental deliverable capacity for Co(bdp) is shown for
       comparison~\cite{fixme}, along with the free energy difference between open and closed conformations~\cite{fixthis}.}
    \label{fig:methane-flexible}
\end{figure}
Although our deliverable capacity limit does not directly apply to materials that change conformation in
response to gas adsorption, we can estimate enhancements in the deliverable capacity provided by
such materials. 
For example, consider a material such as Co(bdp) \cite{mason2015methane} that 
can, actuated by the presence of gas, switch between two structural phases: 
an ``open'' phase where gas molecules can fit in the pore space and a 
collapsed, ``closed'' phase in which no gas can fit inside.
Let $\Delta F=F_{\text{open}}-F_{\text{closed}}$ be the Helmholtz free
energy difference between the two phases in the absence of gas.
% David: introduce units of delta G here.
Suppose the closed phase is more stable ($\implies \Delta F >0$), so that upon a reduction in the pressure of the gas, the material will collapse and expel all residual gas to enhance the deliverable capacity compared to rigid materials that will retain some gas at $\pempty$. 
In addition, we assume the open phase hosts a uniform, attractive potential energy field $\V<0$ for gas molecules to adsorb, so that eqn.~\ref{eq:xi_vs_xi0} holds for the open phase of the material.
For an enhancement of the deliverable capacity by this two-phase, flexible material, 
the open phase must be preferred at $\pfull$, with the closed phase preferred at $\pempty$. 
This requires that, neglecting gas-gas interactions:
% David: doesn't this also neglect entropy? wut about the *free energy* of the gas?
% David: I suggest \Delta G_{host} to be clear?
\begin{align}
    V_{\text{open}}|\rho_{\text{open}}(\pfull)\V| &> \Delta F -pV_{\text{open}} \\
    V_{\text{open}}|\rho_{\text{open}}(\pempty)\V| &< \Delta F -pV_{\text{closed}}
\end{align}
% David: I put \V<0, think we should say the sign of \Delta G to be clear too.
% David: specify units on \delta G so clear where V comes from; might be better to absorb V into delta G?
where $\rho_{\text{open}}(p)$ is the density of gas
in the open phase of the material at pressure $p$. We will neglect the $pV$
terms in the Gibbs free energy, under the assumption that the mechanical
work associated with the transition is neglible when compared with the
adsorption energies involved.
I.e., (1, to be open at $\pfull$) the gas-host attractions at $\pfull$ must overcome the free energy penalty to open the structure and 
(2, to be closed at $\pempty$) the free energy penalty to open the structure must be larger than the (weak) gas-host attractions at $\pempty$.
This approximate model of the two-phase material has no upper bound on the deliverable capacity, since
$\V$ and $\Delta F$ can together be increased without bound.
We can, however, place a bound for a fixed $\Delta F$, since if the gas is too strongly attracted to the open phase,
the material will remain open at $\pempty$ and our rigid upper bound applies.
% David: does our rigid upper bound apply then?
The upper bound for the two-phase material is found by identifying the
attractive interaction $\V$ that balances the gas-host attractions in the idealized-open-substrate at $\pempty$ with the free energy penalty to open:
\begin{align}
    V_{\text{open}} \rho_g(\muempty{} - \V_{opt}, T)\V_{opt} = \Delta F.
\end{align}
% David: left is negative. what is the right? negative as well? how did u define it?
This gives the largest possible attraction $|\V|$ consistent with the material closing
at $\pempty$ to expel all residual gas. 
The deliverable capacity is then the density of gas in the open phase at $\pfull$, $\rho_g(\mufull{} - \V_{opt}, T)$.
% David: changed it to \V_opt, or not necessary?

Figure~\ref{fig:methane-flexible} illustrates
how the upper bound on the deliverable capacity of methane in a two-phase,
collapsing material depends on $\Delta F$. 
For comparison, we include the free energy
difference for Co(bdp)~\davidsays{CITE} and its measured deliverable capacity.  
While this upper bound is not rigorous, it suggests that a two-phase flexible material for methane storage
would benefit from significantly greater stability of the closed phase---provided the open phase
had correspondingly greater attraction for the gas molecules.  Figure~\ref{fig:hydrogen-flexible}
shows the corresponding result for hydrogen storage, indicating that, for a flexible material to
reach the DOE ULTIMATE deliverable capacity, it would require far greater stability for the closed
phase, as well as an unprecedented level of attractive interaction for hydrogen.

@@

\donereferees

@@

\startreferees{Added an explanation of why we cannot place a bound on the
  deliverable capacity when the temperature is changed.}

The first limitation of our proof is that we restricted ourselves to
\emph{isothermal} pressure swing storage (which is consistent with the targets set by DOE). By raising the temperature of the
adsorbent during gas discharge to drive off residual
gas~\cite{gomez2014exploring}, the deliverable capacity could be enhanced,
albeit at the cost of a more complicated engineering design of the fuel tank
and vehicle.
%
We cannot place an upper bound on such a process using our approach because with
changing temperature a uniform potential may not give the maximum deliverable
capacity.  A nonuniform potential will lower the entropy of the adsorbed gas,
which will make the gas easier to release by raising the temperature.

\donereferees

>>>>

\usepackage[color]{changebar}
\cbcolor{red}

@@

\newcommand{\davidsays}[1]{{\color{red} [\blue{D:} \emph{#1}]}}
\newcommand{\jpsays}[1]{{\color{red} [\blue{Jordan:} \emph{#1}]}}

@@

For a vehicle employing a porous material to store natural gas or hydrogen, the
(volumetric) \emph{deliverable capacity} of the gas is the primary
thermodynamic property that determines the driving
range on a given ``full'' tank of fuel~\cite{mason2014evaluating}. 
The adsorbed gas storage tank delivers the
gaseous fuel to the engine via an (assumed) isothermal pressure
swing~\cite{sircar2002pressure}. The deliverable capacity (see
Fig.~\ref{fig:delcap}) is the density of the gas in the material at the storage
pressure $\pfull$ minus the residual gas that remains adsorbed at the lowest
pressure $\pempty$ such that sufficient flow is maintained to feed the engine.
For commercial feasibility, the US Department of Energy (DOE) has set
deliverable capacity targets for both adsorbed methane and hydrogen storage and
delivery for vehicles. For methane, the Advanced Research Projects
Agency--Energy (ARPA-E) set a deliverable capacity target of 315\ L STP CH$_4$
per L of adsorbent at 298 K using a 65\,bar to 5.8\,bar pressure
swing~\cite{simon2015materials}. For hydrogen, the DOE set a series of
progressive targets at five year intervals, with the ultimate target of 0.05\,kg
H$_2$/L~\cite{h2targetsDOE} using a 100\ bar to 5\ bar pressure swing at a
minimum of -40 $^\circ$C~\cite{allendorf2018assessment}. Thus far, despite the
emergence of highly tunable materials with large surface areas, such as
metal-organic frameworks~\cite{furukawa2013chemistry} and covalent organic 
frameworks~\cite{diercks2017atom}, no porous material has met these deliverable capacity targets.

@@

Finally, empirical models that statistically relate e.g., surface area to 
hydrogen adsorption \cite{firlej2013understanding,goldsmith2013theoretical}
have been used to determine what properties a material must possess to reach
a given storage target.

@@

\subsection{The real-substrate analog of $\V$ is $\gst$}
\label{sec:phi-is-delta-g}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{g_st_simple.pdf}
    \caption{A illustrative thought experiment to measure $\gst$ by adjusting the altitute difference between two connected volumes, one free space and another containing a porous material.
    $\gst$ is the difference in the gravitational potential energy when the density of gas in each volume is equal.
    % For a given adsorbent and gas, $\gst$=$\gst (\rho, T)$ is defined as the difference in chemical potential between the adsorbed gas system and bulk gas system when both exhibit the same density $\rho$ and temperature $T$. 
    }
    \label{fig:delta-G-cartoon}
\end{figure}

@@

\begin{equation}
   \gst(\rho, T) \equiv
   %\frac{G_{\text{ads}}(\rho, T) - G_{\text{gas}}(\rho, T)}{N}\\
    \mu_{\text{ads}}(\rho, T) - \mu_{\text{gas}}(\rho, T).
  % \gst(\rho, T) &\equiv g_{\text{gas}}(\rho, T) - g_{\text{ads}}(\rho, T)
  \label{eq:g_st}
\end{equation}
The isosteric Gibbs free energy difference $\gst$ is the difference in molar
Gibbs free energy (equivalent to chemical potential) between the adsorbed gas system and
the bulk gas \emph{with the same density of gas molecules} and at the same temperature. The quantity $\gst$
does \emph{not} correspond to a change in the molar Gibbs free energy as a
molecule is adsorbed, which is zero under conditions of coexistence. The
quantity $\gst$ in a real substrate is a direct analog to $\V$ in our ideal
substrate because it is the chemical potential shift needed to impose on the
bulk gas in coexistence with the real substrate to achieve the same density as
in the substrate (compare with eqn.~\ref{eq:xi_vs_xi0}).
Figure~\ref{fig:delta-G-cartoon} illustrates $\gst$.
% David:  no longer illustrates the expt.
A hypothetical experiment to measure $\gst$ is via a piston with a removable partition that separates a volume of free space from the same volume of substrate.
Initialize the experiment with all gas in the free space and an empty substrate.
Then, slowly (i) remove the partition and (ii) use the piston to push the gas into the substrate and eliminate the free space volume, in an isothermal process.
$\gst$ is the ... \davidsays{wut? energy it takes to do this or?}
% David: what is GST
Note that $\gst$ is a property
of both the substrate and the identity of the gas. Because real substrates
offer a \emph{non}-spatially uniform potential, $\gst(\rho, T)$ is a function
of $\rho$ and $T$, unlike our ideal, homogenous substrate where $\gst(\rho,
T)=\V$. Consequently, throughout this article, we show $\gst(\rho, T)$ for real
substrates at both conditions relevant to gas storage and delivery, $\pfull$
and $\pempty$.

@@

\cbstart
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{figs/two-phase}
    \caption{Cartoon of a two-phase, collapsing material. The material can exist in two phases: an open phase in which gas molecules can adsorb and a closed phase that is collapsed and cannot fit any gas molecules. Take the closed phase as more stable, so that the material can collapse and expels residual gas.
    The pressure of gas at which the structural transition occurs depends on the gas-host interactions and relative stability of the phases. We wish for the structural transition to occur between $\pfull$ and $\pempty$.
    }
    \label{fig:flexible-cartoon}
\end{figure}

\section{Flexible two-phase model}
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{methane-298-flexible}
    \caption{Maximal possible deliverable capacity of methane in a two-phase, collapsing substrate at 298\ K as a function of the free energy difference 
      between open and closed phases. The experimental deliverable capacity for Co(bdp) is shown for
       comparison~\cite{fixme}, along with the free energy difference between open and closed conformations~\cite{fixthis}.}
    \label{fig:methane-flexible}
\end{figure}
Although our deliverable capacity limit does not directly apply to materials that change conformation in
response to gas adsorption, we can estimate enhancements in the deliverable capacity provided by
such materials. 
For example, consider a material such as Co(bdp) \cite{mason2015methane} that 
can, actuated by the presence of gas, switch between two structural phases: 
an ``open'' phase where gas molecules can fit in the pore space and a 
collapsed, ``closed'' phase in which no gas can fit inside.
Let $\Delta G=G_{\text{open}}-G_{\text{closed}}$ be the free energy difference between the two phases in the absence of gas.
% David: introduce units of delta G here.
Suppose the closed phase is more stable ($\implies \Delta G >0$), so that upon a reduction in the pressure of the gas, the material will collapse and expel all residual gas to enhance the deliverable capacity compared to rigid materials that will retain some gas at $\pempty$. 
In addition, we assume the open phase hosts a uniform, attractive potential energy field $\V<0$ for gas molecules to adsorb, so that eqn.~\ref{eq:xi_vs_xi0} holds for the open phase of the material.
For an enhancement of the deliverable capacity by this two-phase, flexible material, 
the open phase must be preferred at $\pfull$, with the closed phase preferred at $\pempty$. 
This requires that, neglecting gas-gas interactions:
% David: doesn't this also neglect entropy? wut about the *free energy* of the gas?
% David: I suggest \Delta G_{host} to be clear?
\begin{align}
    V_{\text{open}}|\rho_{\text{open}}(\pfull)\V| &> \Delta G \\
    V_{\text{open}}|\rho_{\text{open}}(\pempty)\V| &< \Delta G
\end{align}
% David: I put \V<0, think we should say the sign of \Delta G to be clear too.
% David: specify units on \delta G so clear where V comes from; might be better to absorb V into delta G?
where $\rho_{\text{open}}(p)$ is the density of gas in the open phase of the material at pressure $p$. 
I.e., (1, to be open at $\pfull$) the gas-host attractions at $\pfull$ must overcome the free energy penalty to open the structure and 
(2, to be closed at $\pempty$) the free energy penalty to open the structure must be larger than the (weak) gas-host attractions at $\pempty$.
This approximate model of the two-phase material has no upper bound on the deliverable capacity, since
$\V$ and $\Delta G$ can together be increased without bound.
We can, however, place a bound for a fixed $\Delta G$, since if the gas is too strongly attracted to the open phase,
the material will remain open at $\pempty$ and our rigid upper bound applies.
% David: does our rigid upper bound apply then?
The upper bound for the two-phase material is found by identifying the
attractive interaction $\V$ that balances the gas-host attractions in the idealized-open-substrate at $\pempty$ with the free energy penalty to open:
\begin{align}
    V_{\text{open}} \rho_g(\muempty{} - \V_{opt}, T)\V_{opt} = \Delta G.
\end{align}
% David: left is negative. what is the right? negative as well? how did u define it?
This gives the largest possible attraction $|\V|$ consistent with the material closing
at $\pempty$ to expel all residual gas. 
The deliverable capacity is then the density of gas in the open phase at $\pfull$, $\rho_g(\mufull{} - \V_{opt}, T)$.
% David: changed it to \V_opt, or not necessary?

Figure~\ref{fig:methane-flexible} illustrates
how the upper bound on the deliverable capacity in a two-phase, collapsing material depends on $\Delta G$ for methane. 
For comparison, we include the free energy
difference for Co(bdp)~\davidsays{CITE} and its measured deliverable capacity.  
While this upper bound is not rigorous, it suggests that a two-phase flexible material for methane storage
would benefit from significantly greater stability of the closed phase---provided the open phase
had correspondingly greater attraction for the gas molecules.  Figure~\ref{fig:hydrogen-flexible}
shows the corresponding result for hydrogen storage, indicating that, for a flexible material to
reach the DOE ULTIMATE deliverable capacity, it would require far greater stability for the closed
phase, as well as an unprecedented level of attractive interaction for hydrogen.

@@

\cbend

@@

The first limitation of our proof is that we restricted ourselves to
\emph{isothermal} pressure swing storage (which is consistent with the targets set by DOE). By raising the temperature of the
adsorbent during gas discharge to drive off residual
gas~\cite{gomez2014exploring}, the deliverable capacity could be enhanced,
albeit at the cost of a more complicated engineering design of the fuel tank
and vehicle.
